---
title: "Estimating NDs and RFDs"
author: "Zena Lapp"
date: "1/24/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

__Goal:__ Estimate niche differences (NDs) and relative fitness differences (RFDs) for coculture populations of _Eschericia coli_ and _Pseudomonas putida_ growing at different temperatures. 

__Contents:__

1. Methods
    1. Raw data processing
    1. Curve fitting
1. Plots for both monoculture & coculture of:
    1. ODs 
    1. Raw _E. coli_ fluorescence
    1. Raw _P. putida_ fluorescence 
    1. Post-normalization _E. coli_ fluorescence
    1. Post-normalization _P. putida_ fluorescence 
    1. Processed data & fitted data
1. Plot of ND vs. RFD

```{r}
# load libraries & functions
library(ggpmisc)
source('lib/process_raw_data.R')
source('lib/comp_coef_fns.R')
```

## Methods

### Raw data processing

1. __Adjust stationary phase:__ Fluorescence keeps increasing even after the bacteria stop growing. To adjust for this, at each temperature, and for monocultures & cocultures, we:
    1. Manually decide on a maximum value for fluorescence at stationary phase based on where the OD flattens out (suggesting that the bacteria have stopped growing).
    1. Identify the time at which the fluorescence reaches the chosen maximum value and make all fluorescence values after that timepoint the chosen maximum value.
1. __Normalize data:__ The _E. coli_ fluorescence molecule and the _P. putida_ fluorescence molecule have very different frequencies, and the frequencies are different at different temperatures. To adjust for this, we use the monoculture ODs and fluorescences: 
    1. Find the maximum OD and the maximum fluorescence for the _E. coli_ monoculture and for the _P. putida_ monoculture. These are denoted as $E_{OD_{max}}$, $E_{f_{max}}$, $PP_{OD_{max}}$, and $PP_{f_{max}}$, respectively.
    1. Normalize all data to 32&deg;C using $E_{OD_{32}}$, $E_{f_{32}}$, and $P_{OD_{32}}$.
    1. Normalize all _E. coli_ values to: $(E_{f_{32}}/E_{OD_{32}})*(E_{OD_{max}}/E_{f_{max}})$
    1. Normalize all _P. putida_ values to: $(E_{f_{32}}/E_{OD_{32}})*(P_{OD_{max}}/P_{f_{max}})$
    1. Divide all normalized values by 10,000 to improve fit. 

### Curve fitting




Model:

$$\dot{E} = r_EE(1-\alpha_{EE}*E-\alpha_{EP}P)$$
$$\dot{P} = r_PP(1-\alpha_{PP}*P-\alpha_{PE}E)$$

- $E$ is the _E. coli_ population size.
- $P$ is the _P. putida_ population size.
- $r_i$ is the monoculture growth rate.
- $\alpha_{EE}=1/K_E$ and $\alpha_{PP}=1/K_P$ are the intra-specific competition coefficients.
- $K_i$ is the monoculture carrying capacity.
- $\alpha_{EP}$ and $\alpha_{PE}$ are the inter-specific competition coefficients.

We estimate the parameters by fitting the model to the data:

1. $r_i$ and $K_i$ are estimated from monoculture data using the the single-species Lotka-Voltera equations (no $\alpha_{EP}P$ and $\alpha_{PE}E$ terms) using a built-in R function (`growthcurver::SummarizeGrowth`). 
1. $\alpha_{EP}$ and $\alpha_{PE}$ are estimated using the equations above (two-species equation).
    1. Solve the ode for some initial guess of model parameters (`deSolve::ode` function).
        1. Starting values of $\alpha_{EP}$ and $\alpha_{PE}$ for model fitting were obtained by using the steady-state estimates:
        $$\alpha_{EP}=(1-\alpha_{PP}P_{SS})/E_{SS}$$
        $$\alpha_{PE}=(1-\alpha_{EE}E_{SS})/P_{SS}$$
            1. $E_{SS}$ and $P_{SS}$ are estimated using the single-species Lotka Voltera model.
    1. Evaluate an error function between the model results and the data.
        1. Error function:
        $$error=((E_{real}-E_{fitted})/E_{real})^2+((P_{real}-P_{fitted})/P_{real})^2$$
    1. Employ an optimization scheme (`optim` function) to minimize the error function by choosing the set of model parameters that minimizes the error.

## Results

### Raw & normalized data

```{r}
# read in raw data
raw_dat = read_raw_data(dir = 'data/raw/')
# load data about where to truncate (for start of exponential phase) and flatten (start of stationary phase)
trunc_flatten = read.csv('data/trunc_flatten.csv')
# flatten stationary
flattened = flatten_stationary(raw_dat,trunc_flatten)
# normalize data
normalized = normalize_fluorescence(flattened)
# plot raw data
plot_plots(raw_dat,normalized)
```
```{r}
temps = c('32C','34C','36C','38C','40C')
for(i in temps){
  write_csv(x = normalized[[i]], path = paste0('data/normalized/',i,'.csv'))
}
```


### Fitted data

```{r}
# fit data
# initialize dataframe
temps = c('32C','34C','36C','38C','40C')
p_names = c('temp','ec_r','pp_r','ec_k','pp_k','a_ii_start','a_ii','a_jj_start','a_jj','a_ij_start','a_ij','a_ji_start','a_ji','nd','rfd','error')
p = data.frame(matrix(NA, nrow=length(temps),ncol=length(p_names)))
rownames(p) = temps
colnames(p) = p_names

#k_vals = sapply(normalized, function(x) x[nrow(x),c('EC1','PP1')])

plots = list()

# for each temperature
for(temp in temps){
  
  #if(temp == '36C') next()
  
  # get all data for temperature
  d = normalized[[temp]] %>% gather(Replicate,Growth,EC1:EC3,PP1:PP3,ECco1:ECco3,PPco1:PPco3)

  # get EC monoculture data
  ec = filter(d,Replicate %in% c('EC1','EC2','EC3'))
  # get PP monoculture data
  pp = filter(d,Replicate %in% c('PP1','PP2','PP3'))
  # get ECco data
  ecco = filter(d,Replicate %in% c('ECco1','ECco2','ECco3'))
  # get PPco coculture data
  ppco = filter(d,Replicate %in% c('PPco1','PPco2','PPco3'))
  # get all co-culture data
  co = filter(d,Replicate %in% c('ECco1','ECco2','ECco3','PPco1','PPco2','PPco3'))
  
  dat = co

  # get EC parameters
  ec_k = est_k(ec$Time,ec$Growth) #ec$Growth[nrow(ec)]
  ec_r = est_r(ec$Time,ec$Growth)
  # get PP parameters
  pp_k = est_k(pp$Time,pp$Growth) #pp$Growth[nrow(pp)]
  pp_r = est_r(pp$Time,pp$Growth)
  # get ECco parameters
  ecco_k = est_k(ecco$Time,ecco$Growth)
  ecco_n0 = est_n0(ecco$Time,ecco$Growth)
  # get PPco parameters
  ppco_k = est_k(ppco$Time,ppco$Growth)
  ppco_n0 = est_n0(ppco$Time,ppco$Growth)
  
  # print(temp)
  # print('EC')
  # print(ec_k - est_k(ec$Time,ec$Growth))
  # print('PP')
  # print(pp_k - est_k(pp$Time,pp$Growth))

  # # get intra-species competition coefficient (inverse of monoculture carrying capacity)
  # a_ii_start = 1/ec_k
  # a_jj_start = 1/pp_k
  # # get estimate of inter-species competiton coefficient
  # a_ij_start = (1-a_ii_start*ecco_k)/ppco_k
  # a_ji_start = (1-a_jj_start*ppco_k)/ecco_k
  
  
  #if(temp %in% c('32C','34C','40C')) 
  a_ii_start = a_jj_start = a_ij_start = a_ji_start = 0.5
  
  
  if(temp == '32C'){
    dat = dat[dat$Time > 6,]
  }
  if(temp == '34C'){
    dat = dat[dat$Time > 3.5,]
  }
  if(temp == '36C'){
    dat = dat[dat$Time > 5,]
  }
  if(temp == '38C'){
    dat = dat[dat$Time > 5,]
  }
    if(temp == '40C'){
    dat = dat[dat$Time > 5,]
  }

  # if(temp == '36C'){
  #   
  # }
  # if(temp == '38C'){
  #   ec_k = 3.6
  #   #pp_k = 
  # }
  #   
  # a_ij_start = 0.3
  # a_ji_start = 0.3
  
  print(temp)

  aij_vals = seq(0,1,0.1)
  aji_vals = seq(0,0.5,0.1)

  # colums are a_ij vals
  # rows are a_ji vals
  errors = sapply(aij_vals,function(x){
    sapply(aji_vals, function(y){
      #print(c(x,y))
      a_ij_start = x
      a_ji_start = y
        fit <- try(optim(par=c(a_ij_start,a_ji_start,ecco_n0,ppco_n0,a_ii_start,a_jj_start),
               fn=ofun,control = list(maxit = 1000),
               lower=c(x,y,0,0,0,0), upper=c(x+10e10,y+10e10,Inf,Inf,Inf,Inf),method='L-BFGS-B'))
        if(inherits(fit,"try-error")) return(NA)
        return(fit$value)
    })
  })
  print(errors)
  
  get_error = function(a_ij_start,a_ji_start){
    fit <- try(optim(par=c(a_ij_start,a_ji_start,ecco_n0,ppco_n0,a_ii_start,a_jj_start),
               fn=ofun,control = list(maxit = 1000),
               lower=c(a_ij_start,a_ji_start,0,0,0,0), upper=c(a_ij_start+10e10,a_ji_start+10e10,Inf,Inf,Inf,Inf),method='L-BFGS-B'))
        if(inherits(fit,"try-error")) return(NA)
        return(fit$value)
  }
  
  params <- expand.grid(aij_vals = seq(0,1,0.1),aji_vals = seq(0,0.5,0.1)) %>% group_by(aij_vals,aji_vals) %>%
    mutate(error = get_error(aij_vals,aji_vals)) %>% ungroup
  
  params %>% ggplot(aes(x=aij_vals,y=aji_vals,z=error)) +
    stat_contour(aes(color=..level..))
  
  next
    
  # get fit
  fit <- optim(par=c(a_ij_start,a_ji_start,ecco_n0,ppco_n0,a_ii_start,a_jj_start),
               fn=ofun,control = list(maxit = 1000),lower=c(0.1,0.1,0,0,0,0), upper=rep(Inf, 6),method='L-BFGS-B')
    
  fitdat <- data.frame(Time=dat$Time,
                       curve=comp.coeff.curve(fit$par[1],fit$par[2],fit$par[3],fit$par[4],fit$par[5],fit$par[6])[,1:2],
                       dat=dat$Growth,Replicate=dat$Replicate)
  
  plots[[temp]] = ggplot(fitdat,aes(x=Time,y=dat)) +
    geom_point(aes(col=Replicate))+geom_line(aes(y=curve.E)) +
    geom_line(aes(y=curve.P)) + ggtitle(temp) + theme_bw() + ylab('Growth (r.f.u.)')
 #  
 #  lm_eqn_ec <- function(df){
 #    m <- lm(dat ~ curve.E, df)
 #        eq <- substitute(~~italic(r)^2~"="~r2,
 #         list(r2 = format(summary(m)$r.squared, digits = 2)))
 #    as.character(as.expression(eq))
 #  }
 # 
 #  lm_eqn_pp <- function(df){
 #    m <- lm(dat ~ curve.P, df)
 #        eq <- substitute(~~italic(r)^2~"="~r2,
 #         list(r2 = format(summary(m)$r.squared, digits = 2)))
 #    as.character(as.expression(eq))
 # }
 # 
 #  print(ggplot(fitdat[grep('EC',fitdat$Replicate),],aes(x=curve.E,y=dat)) + geom_point(aes(col=Replicate)) +
 #          geom_smooth(method = "lm") +
 #          geom_text(x = 0.1, y = 0.1, label = lm_eqn_ec(fitdat[grep('EC',fitdat$Replicate),]), parse = TRUE) +
 #          ggtitle(paste(temp,'E. coli coculture')) + ylab('Actual growth (r.f.u.)') + xlab('Fitted growth (r.f.u.)') +
 #          theme_bw())
 #  print(ggplot(fitdat[grep('PP',fitdat$Replicate),],aes(x=curve.P,y=dat)) + geom_point(aes(col=Replicate)) +
 #          geom_smooth(method = "lm") +
 #          geom_text(x = 0.1, y = 0.1, label = lm_eqn_pp(fitdat[grep('PP',fitdat$Replicate),]), parse = TRUE) +
 #          ggtitle(paste(temp,'P. putida coculture')) +ylab('Actual growth (r.f.u.)') + xlab('Fitted growth (r.f.u.)') +
 #          theme_bw())
  
  # fitted
  a_ij = fit$par[1]
  a_ji = fit$par[2]
  a_ii = fit$par[5]
  a_jj = fit$par[6]
  
  # calculate ND
  nd = 1 - sqrt((a_ij*a_ji)/(a_ii*a_jj))

  # calculate RFD
  rfd = sqrt((a_ij*a_ii)/(a_ji*a_jj))
  
  # error
  err = fit$value

  # save all to dataframe
  p[temp,] = c(temp, ec_r, pp_r, ec_k, pp_k,a_ii_start, a_ii, a_jj_start, a_jj, 
               a_ij_start, a_ij, a_ji_start, a_ji, nd, rfd, err)
  
}

plot_fitted(plots)
```


### NDs & RFDs

```{r}
p$nd = as.numeric(p$nd)
p$rfd = as.numeric(p$rfd)
rownames(p) = seq(1,length(rownames(p)))
p
p_sub = p#[c(1,2,4,5),]

# plot ND and RFD 
ggplot(data=p_sub)+geom_point(mapping=aes(x=nd,y=rfd,col=temp,size=10))+ylab('rfd')+theme_classic() + stat_function(fun = function(x) 1/(1-x)) +  stat_function(fun = function(x) 1-x)
```


